---
title: "Factors impacting Student Success"
author: "Keelin Joyce"
date: "2026-02-06"
categories: [code, analysis]
---

## Introduction

This dataset looks at 6,607 students and 19 factors that could possibly contribute to exam score. This includes how much students slept and studied, the access to resources they have (including supportive parents and good teachers) and many other variables.

|  |  |  |  |
|----|----|----|----|
| Hours Studied | Attendance | Parental Involvement | Access to Resources |
| Extracurricular Activities | Sleep Hours | Previous Scores | Motivation Level |
| Internet Access | Tutoring Sessions | Family Income | Teacher Quality |
| School Type | Peer Influence | Physical Activity | Learning Disabilites |
| Parental Education_Level | Distance From Home | Gender | Exam Score |

: Variables included in "Student Academic Performance" dataset

<https://www.kaggle.com/datasets/ayeshasiddiqa123/student-perfirmance>

```{r, warning=FALSE, message=FALSE}
library(readr)
library(ggplot2)
library(tidyverse)
StudentPerformanceFactors <- read_csv("~/Desktop/dataviz334/ds334_blog/posts/Factors impacting Student Success/StudentPerformanceFactors.csv")
```

## Question of Interest:

Which factors most directly impact the scores students get on an exam?

## Visualizations

### Exam Score relation to Hours Studied

```{r}
ggplot(data = StudentPerformanceFactors, aes(x = Hours_Studied, y = Exam_Score, colour = Attendance)) +
  geom_point() +
  labs(title = "Relationship between Hours Studied and Exam Score", y = "Exam Score", x = "Hours Studied") +
  geom_smooth(colour = "black") +
  scale_color_viridis_c() +
  theme_minimal()
```

The plot shows the relationship between how many hours a student studied for an exam and how well they scored. It displays a strong, positive correlation that explains that students that studied for longer typically scored higher on an exam. The colouring shows that class attendance and number of hours studied have the same general trend in exam scores. It also shows that there is not an association between attendance and hours studied, meaning that students that study for longer do not necessarily have better class attendance.

### Exam Score relation to Number of Tutoring Sessions

```{r warning=FALSE, echo=FALSE, message=FALSE}
ggplot(data = StudentPerformanceFactors, aes(x = Tutoring_Sessions, y = Exam_Score)) +
  geom_point(colour = "orchid3") +
  labs(title = "Relationship between Number of Tutoring Sessions and Exam Score", y = "Exam Score", x = "Tutoring Sessions") +
  geom_smooth(method = lm, colour = "olivedrab4") +
  theme_minimal()

ggplot(data = StudentPerformanceFactors, aes(x = Tutoring_Sessions)) +
  geom_histogram(fill = "orchid3", colour = "olivedrab4") +
  labs(title = "How Many Tutoring Sessions Students Went to", y = "Count", x = "Tutoring Sessions") +
  theme_minimal()
```

The plot visualizes the relationship between how many tutoring sessions a student attended and how well they scored on the exam. Although there is not a strong positive relationship, there is still generally an increase in Exam Score for students that attend more tutoring sessions. The histogram shows that many went to only 0-3 tutoring sessions and there is not much data for those that went to more.

### Conclusions

Of the many variables observed, only a few have any sort of visible correlation with how students score on an exam. To further analyze this data, statistical tests should be performed to confirm whether there are any associations that can not be determined from visualizations alone.

### Connection to Class Ideas

The graphs demonstrate concepts learned in class of appropriate colour schemes, colouring by variables, and fixing aesthetics to create a better visualization. I utilized concepts from the grammar of graphics unit to enhance the visualizations and make them both easier and more appealing to look at.
